Concentration of Maxima and Fundamental Limits in High-Dimensional Testing and Inference

### 2021-02-01 6.GWAS/GWAS.tex
rewrite GWAS Section intro

We investigate the fundamental limits of multiple testing problems in high-dimensional chi-square models, and in genome-wide association studies as introduced in Section \ref{sec:motivation-chisq}.

In Section \ref{sec:chisq-boundaries}, we shall establish the phase transitions of the sparse chi-square model \eqref{eq:model-chisq}.
Recall that in large-scale screening studies where a large number of association tests are conducted, resulting statistics may be approximated by
\begin{equation} \label{eq:model-chisq-Chapter6}
    %x(i) \distras{\mathrm{ind.}} \chi_\nu^2\left(\lambda(i)\right), \quad i=1,\ldots,p.
    x(i) \sim \chi_\nu^2\left(\lambda(i)\right), \quad i=1,\ldots,p,
\end{equation}
where $\chi_\nu^2\left(\lambda(i)\right)$ is a chi-square distributed random variable with $\nu$ degrees of freedom and non-centrality parameter $\lambda(i)$.
In parallel to results in Chapter \ref{chap:phase-transitions}, we show that several commonly used family-wise error rate-control procedures --- including Bonferroni's procedure --- are asymptotically optimal for the {exact}, and {exact-approximate} support recovery problems (as defined in Definition \ref{def:exact-recovery-success-failure}) in idealized chi-square models \eqref{eq:model-chisq-Chapter6} with independent components.
We further show that the \ac{BH} procedure is asymptotically optimal for the {approximate}, and {approximate-exact} support recovery problems.
% \ref{thm:chi-squared-exact-boundary}, \ref{thm:chi-squared-exact-approx-boundary}, \ref{thm:chi-squared-approx-boundary}, and \ref{thm:chi-squared-approx-exact-boundary}.
Under appropriate parametrizations of the signal sizes and sparsity, they establish the phase transitions of support recovery problems in the chi-square model.
Remarkably, the degree-of-freedom parameter does not affect the asymptotic boundaries in any of the four support recovery problems.

All phase transition boundaries coincide with those in the additive error models obtained in Chapter \ref{chap:phase-transitions} under suitable parametrizations.
% and Figure \ref{fig:phase-chi-squared} continues to apply.
This indicates vanishing differences between the difficulties of the one-sided and two-sided alternatives in the Gaussian additive error model \eqref{eq:model-additive}.

@@

We then return to association screenings of categorical variables in
Section \ref{sec:odds-and-power}, and present the consequences of the phase transition in the exact-approximate problem in large-scale genetic association studies.
% present empirical evidence for the phase transition in the exact-approximate problem using real data from large-scale association studies on breast cancer obtained from the NHGRI-EBI GWAS Catalog \citep{macarthur2016new}.
% demystify the notion of signal size $\lambda$ in this context.
% which is perhaps less transparent than in additive error models.
We do so by characterizing the relationship between the signal size $\lambda$ and the marginal frequencies, odds ratio, and sample sizes for association tests on 2-by-2 contingency tables.
% Specifically, the amount of signal, when rare variants are present, is weaker compared to the signal when the marginal distributions are balanced.
% In other words, reliable detection of the effects by rare variants would require more samples compared to common variants, even at the same odds ratio.
This result, establishing the relationship between sample sizes and signal sizes, is made precise in Section \ref{sec:odds-and-power}.

We elaborate on the implications of this relationship on optimal study designs for association studies in Section \ref{sec:optimal-design}.
Perhaps surprisingly, our analysis reveals that balanced designs with equal number of cases and controls are often statistically inefficient.
Practical consequences of these results in power analysis will be illustrated with data examples in Section \ref{sec:phase-transitions-in-GWAS}.

The phase transitions in the chi-square models are demonstrated with numerical simulations in Section \ref{sec:numerical}.
Proofs of results in this Chapter are collected in Section \ref{sec:proof-signal-size-odds-ratio}.

>>>>

The process of scientific discovery, as famously explained by Richard Faynman, usually starts with guesses in one form or another.
The consequences of such guesses are then computed, and compared with experimental results.
If the predictions disagree with experiment, then our guesses are wrong. ``That is all there is to it''  \citep{feynman2017character}.

In previous chapters, we delved deep in to the theoretical underpinnings of the phase transition phenomena in high-dimensional multiple testing problems.
The results are interesting in their own right.
However, in the spirit of Faynman, we have not discovered any scientific law, but merely worked out mathematical consequences of our postulated models.
In this chapter, our goal is to relate these predictions to real experimental data from the field of genetics, where large-scale simultaneous hypotheses testing problems often arise.
From such comparisons, we will demonstrate that the phase transition laws are indeed accurate predictions of some curious phenomena in the real world.
The accuracy of our predictions will lend credibility to the application of these ``laws of large dimensions'' in actual statistical problems.

@@

In our case, the experimental data used as the measuring stick come from genome-wide association studies (\ac{GWAS}), introduced in Section \ref{sec:motivation-chisq}.
Recall that in \ac{GWAS}, a large number of marginal association tests are conducted simultaneously, resulting in statistics that can be approximated by
\begin{equation} \label{eq:model-chisq-Chapter6}
    %x(i) \distras{\mathrm{ind.}} \chi_\nu^2\left(\lambda(i)\right), \quad i=1,\ldots,p.
    x(i) \sim \chi_\nu^2\left(\lambda(i)\right), \quad i=1,\ldots,p,
\end{equation}
where $\chi_\nu^2\left(\lambda(i)\right)$ is a chi-square distributed random variable with $\nu$ degrees of freedom and non-centrality parameter $\lambda(i)$.

We establish our theoretical predictions in two steps.
In Section \ref{sec:chisq-boundaries} below, we shall first establish the phase transitions of the model \eqref{eq:model-chisq-Chapter6}.
In parallel to results in Chapter \ref{chap:phase-transitions}, we show that several commonly used family-wise error rate-control procedures --- including Bonferroni's procedure --- are asymptotically optimal for the {exact}, and {exact-approximate} support recovery problems (recall Definition \ref{def:exact-recovery-success-failure}) in idealized chi-square models with independent components.
Analogously, the \ac{BH} procedure is asymptotically optimal for the {approximate}, and {approximate-exact} support recovery problems.
% \ref{thm:chi-squared-exact-boundary}, \ref{thm:chi-squared-exact-approx-boundary}, \ref{thm:chi-squared-approx-boundary}, and \ref{thm:chi-squared-approx-exact-boundary}.
Under appropriate parametrizations of the signal sizes and sparsity, they establish the phase transitions of support recovery problems in the chi-square model.
Remarkably, the degree-of-freedom parameter does not affect the asymptotic boundaries in any of the four support recovery problems.

We then translate the canonical signal size and sparsity parametrizations into the vernacular of statistical geneticists in Section \ref{sec:odds-and-power}.
% present empirical evidence for the phase transition in the exact-approximate problem using real data from large-scale association studies on breast cancer obtained from the NHGRI-EBI GWAS Catalog \citep{macarthur2016new}.
% demystify the notion of signal size $\lambda$ in this context.
% which is perhaps less transparent than in additive error models.
We do so by characterizing the relationship between the signal size $\lambda$ and the marginal frequencies, odds ratio, and sample sizes for association tests on 2-by-2 contingency tables. 
These parameters are more often estimated and reported in \ac{GWAS} than the elusive signal size parameter $\lambda$.
% Specifically, the amount of signal, when rare variants are present, is weaker compared to the signal when the marginal distributions are balanced.
% In other words, reliable detection of the effects by rare variants would require more samples compared to common variants, even at the same odds ratio.
% This result, establishing the relationship between sample sizes and signal sizes, is made precise in Section \ref{sec:odds-and-power}.
As a bonus, we point out an implications of this relationship on optimal study designs for association studies in Section \ref{sec:optimal-design}.
Perhaps surprisingly, our analysis reveals that balanced designs with equal number of cases and controls are often statistically inefficient.
% Practical consequences of these results in power analysis will be illustrated with data examples in Section \ref{sec:phase-transitions-in-GWAS}.

Armed with the results on phase transitions in the chi-square model, and a translation from the language of high-dimenstional statistics to the patois of association screening studies, we finally present in Section \ref{sec:phase-transitions-in-GWAS} the consequences of the phase transitions in \ac{GWAS}, and compare against real experimental data to evaluate the success of our predictions.

The phase transitions in the chi-square models are demonstrated with numerical simulations in Section \ref{sec:numerical}.
Proofs are collected in the online Supplement.
%in Section \ref{sec:proof-signal-size-odds-ratio}.
